---
title: "Managing R Package Dependencies via DESCRIPTION"
subtitle: "The R equivalent of requirements.txt in Python"
author: "Harel Lustiger"
date: '2020-03-27'
draft: yes
categories:
- R Projects
- Reproducible Research
slug: managing-r-package-dependencies-via-description
tags:
- Package Management    
- Reproducible Research    
- DESCRIPTION
link-citations: yes 
bibliography: [references-posts.bib]
biblio-style: apalike
---

```{r, include = FALSE}
path_root <- rprojroot::find_root(rprojroot::has_file("DESCRIPTION"))
source(file.path(path_root, "R/post_options.R"))
```

# Synopsis

In This Post

* Introducing an existing package management approach we could use while
developing R extensions such as R packages, R projects and Shiny apps.
* Using a package manager that capitalises on existing infrastructure. 
* Forming a discipline of grooming the `DESCRIPTION` file

The approach in a nutshell:

1. Maintaining package requirements in a `DESCRIPTION` file; 
2. Setting the URL from which to pull the packages from; and
3. Using `remotes::install_deps()` to install the required packages.

Main features:

* Native solution - the DESCRIPTION is a standard file used in any R package. This has two implications:
    1. Easy to use: There is an existing toolkit to maniplualte the DESCRIPTION content.
    2. Cohesive to R package development. Either you are familier with the DESCRIPTION file, or it will facilitate you when you learn to develop R packages.

# Introduction

## The Problem

All R applications have requirements which are necessary for them to run. A
predominant type of requirement is package dependencies. Most R applications are
built upon R packages which do not come with a fresh installation of R. Thus, to
ensure an R application will run on different computers, it is necessary to
manage its package dependencies. There are a few possible package management
strategies to employ. A common thread of these strategies is the installation of
missing R packages.

R workhorse function for installing R packages is `utils::install.packages()`. A
glimpse into the function help documentation shows us what input arguments
impact the function behaviour.

```{r install.packages, echo = TRUE, eval = FALSE}
install.packages(pkgs, lib, repos = getOption("repos"),
                 contriburl = contrib.url(repos, type),
                 method, available = NULL, destdir = NULL,
                 dependencies = NA, type = getOption("pkgType"),
                 configure.args = getOption("configure.args"),
                 configure.vars = getOption("configure.vars"),
                 clean = FALSE, Ncpus = getOption("Ncpus", 1L),
                 verbose = getOption("verbose"),
                 libs_only = FALSE, INSTALL_opts, quiet = FALSE,
                 keep_outputs = FALSE, ...)
```

The first three arguments `pkgs`, `lib` and `repos` specify what to install,
where to install it and where to get it from, respectively. Among different
package management strategies, `pkgs` should be (almost) identical. However, the
methods employed for gathering `pkgs` varies between strategies. Other
noticeable differences between strategies, are the values passed to `lib` and
`repos`. The following [recipe](#Recipe) elaborates how to curate `pkgs` and how
to leverage `repos` to emulate CRAN (the default package collection) at a
particular date.


<!-- The objectives of the strategy in this document are -->
<!-- https://www.infoq.com/presentations/Simple-Made-Easy/ -->

# Recipe {#Recipe}

## Ingredients

* The [`remotes`](https://github.com/r-lib/remotes) package;
* The [`usethis`](https://github.com/r-lib/usethis/) package; and
* A `DESCRIPTION` file.

You can install both packages at once by running:

```{r, echo = TRUE, eval = FALSE}
install.packages(c("remotes", "usethis"))
```

You can create a new `DESCRIPTION` file by running:

```{r, echo = TRUE, eval = FALSE}
usethis::use_description()
```

The last command saves a `DESCRIPTION` skeleton at the working directory with
the following content:

```
Package: project.name
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: 
    person(given = "First",
           family = "Last",
           role = c("aut", "cre"),
           email = "first.last@example.com",
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: What the package does (one paragraph).
License: What license it uses
Encoding: UTF-8
LazyData: true
```

`DESCRIPTION` uses an archaic file format called DCF (Debian control format)
which was available during R inception. DCF is somewhat similar to YAML in its
structure. Each line consists of a **field** name paired with one or more
**values** separated by a colon. When values span multiple lines, they need to
be indented by four spaces.

Package management in R is done primarily through assigning packages names to
the `Imports` field which we handle next.

## Directions

### Grooming the `DESCRIPTION` file

As the R application evolves, anytime an additional package is required for the
application to work, specify that packages in the `DESCRIPTION` file. You can
add a package to the `DESCRIPTION` file by running:

```{r, echo = TRUE, eval = FALSE}
usethis::use_package("dplyr")
```

Alternatively, you can declare the package name and minimum version by running:

```{r, echo = TRUE, eval = FALSE}
usethis::use_package("dplyr", min_version = "0.8.0")
```

The last command appends `dplyr (>= 0.8.0)` under `Imports` in `DESCRIPTION`:

```
Package: project.name
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: 
    person(given = "First",
           family = "Last",
           role = c("aut", "cre"),
           email = "first.last@example.com",
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: What the package does (one paragraph).
License: What license it uses
Encoding: UTF-8
LazyData: true
Imports: 
    dplyr (>= 0.8.0)
```

`usethis::use_package` takes care of syntax, duplication and typos (both in
package name and version) for you.

Removing a package from the `DESCRIPTION` file is done manually. Open
`DESCRIPTION` and delete the package name from the `Imports` fields. Make sure
the last package in `Imports` doesn't end with a comma (',').

### Installing Package Requirements

Firstly, replicate the package versions that were available at the time of
developing the program. Following this step increases application behaviour
consistency regardless of the date it is executed.

(recommended) You can specify `repos` to point at a snapshot in time of CRAN by
running:

```{r, echo = TRUE, eval = FALSE}
options(repos = "https://mran.microsoft.com/snapshot/2020-03-27")
```

(replace _2020-03-27_ with the date required for reproducibility)

If you skip that step, R installs the latest package versions on the hosting
computer.

Lastly, Given an R project with a `DESCRIPTION` file in the root folder, when
`remotes::install_deps()` is called, then R tries to install the listed packages
under `Imports` within `DESCRIPTION`.

The following is an extract from the output console:

```
> remotes::install_deps()

Installing packages into '/usr/local/lib/R/site-library'
(as 'lib' is unspecified)

trying URL 'https://mran.microsoft.com/snapshot/2020-03-27/src/contrib/dplyr_0.8.5.tar.gz'
Content type 'application/octet-stream' length 3229510 bytes (3.1 MB)
==================================================
downloaded 3.1 MB
```

Notice that `remotes::install_deps`:

* Searches and pulls the `dplyr` package from the `repos` address we have
provided it; and

* Installs `dplyr` version 0.8.5. 
    * While we declared the minimum version to be 0.8.0 in `DESCRIPTION`, the
    available version on 2020-03-27 is 0.8.5.
    * `remotes::install_deps` fails if any declared minimum version is not
    available; otherwise, it installs the latest available version on `repos`.


# Installing Packages Not on CRAN 

There are two good reasons why one may want to override the installation of
packages from the URL defined in `repos`:

1. A particular R package is not available in `repos`, but is available
somewhere else; and
2. A particular R package is available in `repos`, but a specific version is
required.

To install `dplyr` version 0.8.0, despite the availability of a newer version
(0.8.5) on `repos`, we add the `Remotes` field in the `DESCRIPTION` file. 
The remote dependencies specified in Remotes should be described in the following form:

```
Remotes: 
    [type::]<Repository>
```

The `dplyr` package is available on GitHub under `tidyverse/dplyr` repository,
i.e. <https://github.com/tidyverse/dplyr>. Thus, in this case, *type = github*
and *Repository = tidyverse/dplyr*. The DESCRIPTION file becomes:

```
Package: project.name
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: 
    person(given = "First",
           family = "Last",
           role = c("aut", "cre"),
           email = "first.last@example.com",
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: What the package does (one paragraph).
License: What license it uses
Encoding: UTF-8
LazyData: true
Imports: 
    dplyr (>= 0.8.0)
Remotes:
    github::tidyverse/dplyr
```

Without specifying more information, `remotes::install_deps` will pull and
install the latest version available in the GitHub Repo (on the master branch).
Instead, we can define a specif git commit to install:

* Install specific release: [`github::tidyverse/dplyr@v0.8.0`](https://github.com/tidyverse/dplyr/releases/tag/v0.8.0)
```
Remotes:
    github::tidyverse/dplyr@v0.8.0
```
* Install specific commit: [`github::tidyverse/dplyr@f05230dd`](https://github.com/tidyverse/dplyr/tree/f05230dd110a729e95b9ffb86df2030bf3b8b26b)
```
Remotes:
    github::tidyverse/dplyr@f05230dd
```

# Conclusion

Putting it all together, we get:

```{r conclusion, eval = FALSE, echo = TRUE}
# During project development use this code --------------------------------
## Set the repos link for a specific date
options(repos = "https://mran.microsoft.com/snapshot/2020-03-27")

## Install the required packages
install.packages(c("remotes", "usethis"))

## Create a DESCRIPTION file
usethis::use_description()

## Add package dependencies to the DESCRIPTION file
usethis::use_package("dplyr", min_version = "0.8.0")
usethis::use_package("ggplot")

# During project deployment use this code ---------------------------------
## Set the repos link for a specific date
options(repos = "https://mran.microsoft.com/snapshot/2020-03-27")
## Install all package dependencies
remotes::install_deps()
```

You could try it yourself on RStudio cloud by [clicking here](https://rstudio.cloud/project/1179635)

# Further Reading

* [Install R Packages from GitHub, BitBucket, or other local or remote
repositories](https://remotes.r-lib.org/articles/dependencies.html)
