---
title: "R Package Management via DESCRIPTION"
author: "Harel Lustiger"
date: '2020-03-27'
output: word_document
bibliography: references-posts.bib
draft: yes
link-citations: yes
categories:
- R Projects
- Reproducible R
slug: r-package-management-via-description
tags:
- package management
- DESCRIPTION
biblio-style: apalike
---

```{r, include = FALSE}
path_root <- rprojroot::find_root(rprojroot::has_file("DESCRIPTION"))
source(file.path(path_root, "R/post_options.R"))
```

## Synopsis

**In This Post**

* Introducing an existing package management approach we could use while
developing R extensions such as R packages, R projects and Shiny apps.
* Using a package manager that capitalises on existing infrastructure. 
* Forming a discipline of grooming the `DESCRIPTION` file

The approach in a nutshell:

1. Maintaining package requirements in a `DESCRIPTION` file; 
2. Setting the URL from which to pull the packages from; and
3. Using `remotes::install_deps()` to install the required packages.

## Introduction

All R applications have requirements which are necessary for them to run. A
predominant type of requirement is package dependencies. Most R applications are
built upon R packages which do not come with a fresh installation of R. Thus, to
ensure an R application will run on different computers, it is necessary to
manage its package dependencies. There are a few possible package management
strategies to employ. A common thread of these strategies is the installation of
missing R packages.

R workhorse function for installing R packages is `utils::install.packages()`. A
glimpse into the function help documentation shows us what input arguments
impact the function behaviour.

```{r install.packages, echo = TRUE, eval = FALSE}
install.packages(pkgs, lib, repos = getOption("repos"),
                 contriburl = contrib.url(repos, type),
                 method, available = NULL, destdir = NULL,
                 dependencies = NA, type = getOption("pkgType"),
                 configure.args = getOption("configure.args"),
                 configure.vars = getOption("configure.vars"),
                 clean = FALSE, Ncpus = getOption("Ncpus", 1L),
                 verbose = getOption("verbose"),
                 libs_only = FALSE, INSTALL_opts, quiet = FALSE,
                 keep_outputs = FALSE, ...)
```

The first three arguments `pkgs`, `lib` and `repos` specify what to install,
where to install it and where to get it from, respectively. Among different
package management strategies, `pkgs` should be (almost) identical. However, the
methods employed for gathering `pkgs` varies between strategies. Other
noticeable differences between strategies, are the values passed to `lib` and
`repos`. The following [recipe](#Recipe) elaborates how to curate `pkgs` and how
to leverage `repos` to emulate CRAN (the default package collection) at a
particular date.

<!-- The objectives of the strategy in this document are -->
<!-- https://www.infoq.com/presentations/Simple-Made-Easy/ -->

## Recipe {#Recipe}

### Ingredients

* The [`remotes`](https://github.com/r-lib/remotes) package;
* The [`usethis`](https://github.com/r-lib/usethis/) package; and
* A `DESCRIPTION` file.

You can install both packages at once by running:

```{r, echo = TRUE, eval = FALSE}
install.packages(c("remotes", "usethis"))
```

You can create a new `DESCRIPTION` file by running:

```{r, echo = TRUE, eval = FALSE}
usethis::use_description()
```

The last command saves a `DESCRIPTION` skeleton at the working directory with
the following content:

```
Package: project.name
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: 
    person(given = "First",
           family = "Last",
           role = c("aut", "cre"),
           email = "first.last@example.com",
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: What the package does (one paragraph).
License: What license it uses
Encoding: UTF-8
LazyData: true
```

### Directions

#### Grooming the `DESCRIPTION` file

As the R application evolves, anytime an additional package is required for the
application to work, specify that packages in the `DESCRIPTION` file. You can
add a package to the `DESCRIPTION` file by running:

```{r, echo = TRUE, eval = FALSE}
usethis::use_package("dplyr")
```

Alternatively, you can declare the package name and minimum version by running:

```{r, echo = TRUE, eval = FALSE}
usethis::use_package("dplyr", min_version = "0.8.0")
```

The last command appends `dplyr (>= 0.8.0)` under `Imports` in `DESCRIPTION`:

```
Package: project.name
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: 
    person(given = "First",
           family = "Last",
           role = c("aut", "cre"),
           email = "first.last@example.com",
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: What the package does (one paragraph).
License: What license it uses
Encoding: UTF-8
LazyData: true
Imports: 
    dplyr (>= 0.8.0)
```

`usethis::use_package` takes care of syntax, duplication and typos (both in
package name and version) for you.

Removing a package from the `DESCRIPTION` file is done manually. Open
`DESCRIPTION` and delete the package name from the `Imports` fields. Make sure
the last package in `Imports` doesn’t end with a comma (',').

#### Installing Package Requirements

Firstly, replicate the package versions that were available at the time of
developing the program. Following this step increases application behaviour
consistency regardless of the date it is executed.

(recommended) You can specify `repos` to point at a snapshot in time of CRAN by
running:

```{r, echo = TRUE, eval = FALSE}
options(repos = "https://mran.microsoft.com/snapshot/2020-03-27")
```

(replace _2020-03-27_ with the date required for reproducibility)

If you skip that step, R installs the latest package versions on the hosting
computer.

Lastly, Given an R project with a `DESCRIPTION` file in the root folder, when
`remotes::install_deps()` is called, then R tries to install the listed packages
under `Imports` within `DESCRIPTION`.

The following is an extract from the output console:

```
> remotes::install_deps()

Installing packages into ‘/usr/local/lib/R/site-library’
(as ‘lib’ is unspecified)

trying URL 'https://mran.microsoft.com/snapshot/2020-03-27/src/contrib/dplyr_0.8.5.tar.gz'
Content type 'application/octet-stream' length 3229510 bytes (3.1 MB)
==================================================
downloaded 3.1 MB
```

Notice that `remotes::install_deps`:

* Searches and pulls the `dplyr` package from the `repos` address we have
provided it; and

* Installs `dplyr` version 0.8.5. 
    * While we declared the minimum version to be 0.8.0 in `DESCRIPTION`, the
    available version on 2020-03-27 is 0.8.5.
    * `remotes::install_deps` fails if any declared minimum version is not
    available; otherwise, it installs the latest available version on `repos`.


## Further Reading

* [Install R Packages from GitHub, BitBucket, or other local or remote
repositories](https://remotes.r-lib.org/articles/dependencies.html)


```
Package: project.name
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: 
    person(given = "First",
           family = "Last",
           role = c("aut", "cre"),
           email = "first.last@example.com",
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: What the package does (one paragraph).
License: What license it uses
Encoding: UTF-8
LazyData: true
Imports: 
    dplyr (>= 0.8.0)
Remotes:
    github::tidyverse/dplyr
```

Other variants

* Install specific release: [`github::tidyverse/dplyr@v0.8.0`](https://github.com/tidyverse/dplyr/releases/tag/v0.8.0)
* Install specific commit: [`github::tidyverse/dplyr@f05230dd`](https://github.com/tidyverse/dplyr/tree/f05230dd110a729e95b9ffb86df2030bf3b8b26b)

